How to Avoid OOTA Without Really Trying

### 2024-10-11 paper/ootavacuous.tex
Fix wording about breaking dependencies
The example of breaking dependencies in 4.3 should be worded better.

Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

In broad terms, OOTA occurs when a group of threads load from each others'
stores and each thread's store depends on the value returned by that
thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air''
as shown in
Listing~\ref{lst:Simple OOTA}.
Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
values of 42, despite there being no instances of 42 in the default-zero
initial values or in the executable
code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
	Throughout this paper, variables whose names are ``\co{r}''
	followed by a number denote private per-thread variables
	(``registers''),
	and other names denote shared variables.
	All shared-variable accesses are relaxed atomic.}
First, thread T0's line~1 does a load from \co{x} into \co{r1},
reading the value of T1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, T0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, T1's line~1 loads 42 from \co{y} to \co{r2}.
Finally, T1's line~2 stores 42 to \co{x}, justifying the value loaded
by T0's line~1.

@@

Inventing atomic loads can also result in OOTA-like behavior.
For example, consider this code:
\begin{quote}
\begin{verbatim}
   int r1 = (x != 0);
   int r2 = (y != 0);
   z = (r1 == r2);
\end{verbatim}
\end{quote}
It is clear that the store to \co{z} semantically depends on the load
from \co{y}, because the value of \co{z} will change whenever \co{y}
changes between zero and nonzero (all else being equal).

@@

Additional future work could:
(1)~Extend these results to some classes of interpreters and JITs on
one hand, and to link-time optimizations (LTO) on the other;
(2)~Consider various classes of global analysis, such as
demonstrating that a given expression will always evaluate to a constant;
(3)~Examine the possibility of ``flattening'' optimizations that combine
multiple threads into one;
(4)~Explore cases in which semantic dependencies must be absolute,
not relative to any particular compiler;
(5)~Include the effects of input and output or other operations which
can vary from one execution to another; and
(6)~Delineate more precisely the limits of permissible behavior for
quasi-volatile object accesses.

\section{Summary and Conclusions}
\label{sec:Summary and Conclusions}

@@

One conclusion to be drawn from this work is that in this context,
avoiding OOTA cycles requires no changes to the standard,
to current implementions, or to user practices for portable code.
This work also demonstrates the need for caution when optimizing
atomic operations.

>>>>

In broad terms, OOTA occurs when a group of threads load from each others'
stores and each thread's store depends on the value returned by that
thread's load.
The collection of loads and stores forms an \emph{OOTA cycle}.
In the most extreme cases a nonsensical value can pop up ``out of thin air''
as shown in
Listing~\ref{lst:Simple OOTA}.
Here, all of \co{x}, \co{y}, \co{r1}, and \co{r2} might have final
values of 42, despite there being no instances of 42 in the default-zero
initial values or in the executable
code~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.\footnote{
	Throughout this paper, variables whose names begin with ``\co{r}''
	denote private per-thread variables (``registers''),
	and other names denote shared variables.
	All shared-variable accesses are relaxed atomic.}
First, thread T0's line~1 does a load from \co{x} into \co{r1},
reading the value of T1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, T0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, T1's line~1 loads 42 from \co{y} to \co{r2}.
Finally, T1's line~2 stores 42 to \co{x}, justifying the value loaded
by T0's line~1.

@@

Inventing atomic loads can also destroy dependencies.
For example, consider this code:
\begin{quote}
\begin{verbatim}
   int r1 = (x != 0);
   int r2 = (y != 0);
   z = (r1 == r2);
\end{verbatim}
\end{quote}
It seems clear that the store to \co{z} semantically depends on the load
from \co{y}, because the value of \co{z} will change whenever \co{y}
changes between zero and nonzero (all else being equal).

@@

Additional future work could:
(1)~Extend these results to some classes of interpreters and JITs on
one hand, and to link-time optimizations (LTO) on the other;
(2)~Consider various classes of global analysis, such as
demonstrating that a given expression will always evaluate to a constant;
(3)~Examine the possibility of ``flattening'' optimizations that combine
multiple threads into one;
(4)~Explore cases in which semantic dependencies must be absolute,
not relative to any particular compiler;
(5)~Include the effects of input and output or other operations which
can vary from one execution to another; and
(6)~Delineate more precisely the limits of permissible behavior for
quasi-volatile object accesses.
This last item may well shed additional light on the dangers of
optimizations involving non-volatile atomic operations beyond the cautions
outlined in
Section~\ref{sec:Inventing Atomic Loads Can Cause Errors and
Destroy Semantic Dependencies}.

\section{Summary and Conclusion}
\label{sec:Summary and Conclusion}

@@

One conclusion to be drawn from this work is that in this context,
avoiding OOTA cycles requires no changes to the standard,
to current implementions, or to user practices for portable code.
% This work also demonstrates the need for caution when optimizing
% atomic operations.
